Random Generation of Rich \\ Abstract Data Type Values

### 2019-01-31 paper/abstract.tex
Pass over abstract
Minor fixes

  % Status
  Automatic generation of random values described by algebraic data types (ADTs)
  is often a hard task.
  % data types (ADTs), % ---specially when considering ADTs which
  % represent HTML pages, images, programs or other structural rich objects.
  % Situation
  State-of-the-art random testing tools can automatically synthesize random data
  generators based on ADTs definitions.
  %
  % In that manner, generated values comply with the structure described by
  % ADTs---something that proves useful when testing software which expects
  % complex input like web pages, images, or even programs.
  %
  In that manner, generated values comply with the structure described by ADTs,
  something that proves useful when testing software which expects complex
  inputs.
  %
  % Problem
  However, it sometimes becomes necessary to generate structural richer ADTs
  values in order to test deeper software layers.
  %
  % Solution
  % In this light, we propose to leverage static information found in the
  In this work we propose to leverage static information found in the codebase
  as a manner to improve the generation process.
  %
  Namely, our generators are capable to consider how programs branch on input
  data as well as how ADTs values are built via interfaces.
  %
  % Concretely, we propose to consider not only ADTs definitions, but also how
  % programs branch on input data as well as how ADTs values get manipulated via
  % interfaces.
  %
  We implement a tool, responsible to synthezise generators for ADTs values
  while providing compile-time guarantees about their distributions.
  %
  Using compile-time predictions, we provide an heuristic that tries to adjust
  the distribution of generators to what developers might want.
  % \todo[inline,author=Ale]{Case study}
  Preliminary experiments of our approach are very encouraging.
\end{abstract}

>>>>

  % Status
  Automatic generation of random values described by algebraic data types (ADTs)
  is often a hard task.
  % data types (ADTs), % ---specially when considering ADTs which
  % represent HTML pages, images, programs or other structural rich objects.
  % Situation
  State-of-the-art random testing tools can automatically synthesize random data
  generators based on ADTs definitions.
  %
  % In that manner, generated values comply with the structure described by
  % ADTs---something that proves useful when testing software which expects
  % complex input like web pages, images, or even programs.
  %
  In that manner, generated values comply with the structure described by ADTs,
  something that proves useful when testing software which expects complex
  inputs.
  %
  % Problem
  However, it sometimes becomes necessary to generate structural richer ADTs
  values in order to test deeper software layers.
  %
  % Solution
  % In this light, we propose to leverage static information found in the
  In this work we propose to leverage static information found in the codebase
  as a manner to improve the generation process.
  %
  Namely, our generators are capable to consider how programs branch on input
  data as well as how ADTs values are built via interfaces.
  %
  % Concretely, we propose to consider not only ADTs definitions, but also how
  % programs branch on input data as well as how ADTs values get manipulated via
  % interfaces.
  %
  We implement a tool, responsible for synthesizing generators for ADTs values
  while providing compile-time guarantees about their distributions.
  %
  Using compile-time predictions, we provide a heuristic that tries to adjust
  the distribution of generators to what developers might want.
  % \todo[inline,author=Ale]{Case study}
  We report on preliminary experiments where our approach shows encouraging
  results.
\end{abstract}

### 2019-01-31 paper/introduction.tex
Minor fixes

% Scenario of random testing
Random testing is a promising approach to finding\todoale{find?} bugs
\cite{HughesNSA16,HughesPAN16,ArtsHNS15}.
%
\quickcheck \cite{ClaessenH00} is the dominant tool of this sort used by the
Haskell community.
%
It requires developers to specify (i) \emph{testing properties} describing
programs expected behavior and (ii) \emph{random data generators} based on the
\emph{types} of the expected inputs (e.g., an integer, an string,
etc.). %, e.g., an integer, an string, or an abstract data type (ADT).
%
\quickcheck then generates random test cases and reports violating testing
properties.

% Situation with Quick Check, ADTs, tools
\quickcheck comes equipped with random generators for built-in types, while it
requires to manually write generators for user-defined ADTs.
%
Recently, there have been a proliferation of tools to automatically derive
\quickcheck generators for ADTs
\cite{mitchell2007,RuncimanNL08,DuregardJW12,grieco2017,DBLP:conf/haskell/MistaRH18}.
%
The main difference about these tools lies on the guarantees provided to ensure
\emph{the termination of the generation process} and the \emph{distribution of
  random values}.
%
Despite their differences, these tools guarantee that generated values are
\emph{well-typed}.
%
In other words, generated values follow the structure described by ADT
definitions.

@@

%% Our proposal
In this work, we identify two different sources of structural information that
can be statically exploited to improve the generation process of ADT values
(Section \ref{sec:sources}).
%
% In this work, we propose a novel improvement in the generation process of ADT
% values by exploiting some static information found in the codebase.
%
Then, we show how to capture this information into our (automatically) derived
random generators.
%
More specifically, we propose a generation process that is capable to consider
how programs branch on input ADTs values as well as how they get manipulated by
abstract interfaces (Section \ref{sec:hrepcont}).
%
Furthermore, we show how to predict the distribution of the \emph{expected}
numbers of ADT constructors, values fitting certain branching patterns, and
calls to interfaces.
%
For that, we extend some recent results on applying \emph{branching
  processes}\cite{gw1875}---a simple stochastic model conceived to study
population growth (Section \ref{sec:synthesis}).
%
%---to predict the distribution of QuickCheck
%generators\cite{DBLP:conf/haskell/MistaRH18}.
%
We implement our ideas in a tool called \dragenp\footnote{\dragenp is available
  at \url{http://github.com/OctopiChalmers/dragen-struct}}, that is capable to
automatically synthesize \quickcheck generators capable to produce \emph{rich ADT values},
where the distributions of random values can be adjusted at compile-time to what
developers might want. %based on our predictions.
%
Finally, we provide empirical evaluations showing that including static
information from the user codebase improves the code coverage of two external
applications when tested using random values generated following our ideas
(Section \ref{sec:casestudies}).

>>>>

% Scenario of random testing
Random testing is a promising approach for finding bugs
\cite{HughesNSA16,HughesPAN16,ArtsHNS15}.
%
\quickcheck \cite{ClaessenH00} is the dominant tool of this sort used by the
Haskell community.
%
It requires developers to specify (i) \emph{testing properties} describing
programs expected behavior and (ii) \emph{random data generators} based on the
\emph{types} of the expected inputs (e.g., an integer, an string,
etc.). %, e.g., an integer, an string, or an abstract data type (ADT).
%
\quickcheck then generates random test cases and reports violating testing
properties.

% Situation with Quick Check, ADTs, tools
\quickcheck comes equipped with random generators for built-in types, while it
requires to manually write generators for user-defined ADTs.
%
Recently, there has been a proliferation of tools to automatically derive
\quickcheck generators for ADTs
\cite{mitchell2007,RuncimanNL08,DuregardJW12,grieco2017,DBLP:conf/haskell/MistaRH18}.
%
The main difference about these tools lies on the guarantees provided to ensure
\emph{the termination of the generation process} and the \emph{distribution of
  random values}.
%
Despite their differences, these tools guarantee that generated values are
\emph{well-typed}.
%
In other words, generated values follow the structure described by ADT
definitions.

@@

%% Our proposal
In this work, we identify two different sources of structural information that
can be statically exploited to improve the generation process of ADT values
(Section \ref{sec:sources}).
%
% In this work, we propose a novel improvement in the generation process of ADT
% values by exploiting some static information found in the codebase.
%
Then, we show how to capture this information into our (automatically) derived
random generators.
%
More specifically, we propose a generation process that is capable of
considering how programs branch on input ADTs values as well as how they get
manipulated by abstract interfaces (Section \ref{sec:hrepcont}).
%
Furthermore, we show how to predict the \emph{expected} distribution of the ADT
constructors, values fitting certain branching patterns, and calls to interfaces
that our random generators produce.
%
For that, we extend some recent results on applying \emph{branching
  processes}\cite{gw1875}---a simple stochastic model conceived to study
population growth (Section \ref{sec:synthesis}).
%
%---to predict the distribution of QuickCheck
%generators\cite{DBLP:conf/haskell/MistaRH18}.
%
We implement our ideas in a tool called \dragenp\footnote{\dragenp is available
  at \url{http://github.com/OctopiChalmers/dragen-struct}}, that is capable of
automatically synthesizing \quickcheck generators which produce \emph{rich ADT
  values}, where the distributions of random values can be adjusted at
compile-time to what developers might want. %based on our predictions.
%
Finally, we provide empirical evaluations showing that including static
information from the user codebase improves the code coverage of two external
applications when tested using random values generated following our ideas
(Section \ref{sec:casestudies}).

### 2019-01-31 paper/synthesis.tex
Minor fixes

Characterizing the distribution of values of an arbitrary random generator is a
hard task.
%
It requires to model every random choice that a generator could possibly make to
generate a value.
%
In a recent work \cite{DBLP:conf/haskell/MistaRH18}, we have shown that it is
possible to \emph{analitically} predict the average distribution of data
constructors produced by a random generators automatically derived considering
only ADT definitions---like the one presented on Section
\ref{sec:randomtesting}.
%
For this purpose, we found that random generation of ADT values can be
characterized using the theory of \emph{branching processes} \cite{gw1875}.
%
This probabilistic theory was originally conceived to predict the growth and
extinction of royal family trees the Victorian Era, being later applied to a
wide variety of research areas.
%
In this work, we adapt this model to also predict the average distribution of
values of random generators derived considering structural information coming
from functions pattern matchings and abstract interfaces.

@@

The red-light colored matrix is what we had before where the light-blue colored
cells are new---we encourage readers to obtain a colored copy of this work.

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of recursive sub-terms of each construction that we
generate (i.e., data constructors, pattern matching values or function calls),
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P})^{n}
\end{align*}
%
The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as shown in Fig. \ref{fig:vectors} (where dashed lines indicate the
composition of the diffent sub-vectors that are calculated separately).

@@

It is possible to provide an overall prediction of the expected number of
constructors when restricting the generation process to only bare data
constructors and pattern matching values.
%
To achieve that, we should stop considering pattern matching values as atomic
values and start seeing them as compositions of several data constructors.
%
In that manner, it is possible to obtain the expected \emph{total} number of
generated data constructors that our generators will produce---regardless if they
are generated on their own, or as part of a pattern matching value.
%
We note this number as $E^\downarrow\![\_]$ and, to calculate it, we only need
to add the expected number of bare constructors that are included within each
pattern matching.
%
For instance, we can calculate the total expected number of constructors |Text|
and |Join| that we will generate by simply expanding the expected number of
generated pattern matching values $simplify\#1$ and $simplify\#2$ into their
corresponding data constructors:
%
{\small
  \begin{align*}
    E^\downarrow\!\!\left[ Text \right]
    = E\left[ Text \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#2 \right] \\
    E^\downarrow\!\!\left[ Join \right]
    = E\left[ Join \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#2 \right]
  \end{align*}
}
%
Observe that each time we generate a value satisfying the first pattern matching
of the function |simplify|, we add two |Text| and one |Join| data constructors
to our random value.
%
The case of the second pattern matching of |simplify| follows analogously.
%
Note that the overall prediction cannot be applied if we also generate random
values containing function calls, as we cannot predict the output of an abstract
function without solving the halting problem in the process.

>>>>

Characterizing the distribution of values of an arbitrary random generator is a
hard task.
%
It requires modeling every random choice that a generator could possibly make to
generate a value.
%
In a recent work \cite{DBLP:conf/haskell/MistaRH18}, we have shown that it is
possible to \emph{analytically} predict the average distribution of data
constructors produced by random generators automatically derived considering
only ADT definitions---like the one presented on Section
\ref{sec:randomtesting}.
%
For this purpose, we found that random generation of ADT values can be
characterized using the theory of \emph{branching processes} \cite{gw1875}.
%
This probabilistic theory was originally conceived to predict the growth and
extinction of royal family trees the Victorian Era, later being applied to a
wide variety of research areas.
%
In this work, we adapt this model to predict the average distribution of values
of random generators derived considering structural information coming from
functions' pattern matchings and abstract interfaces.

@@

The light-red colored matrix is what we had before, whereas the light-blue
colored cells are new---we encourage readers to obtain a colored copy of this
work.

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of recursive sub-terms of each construction that we
generate (i.e., data constructors, pattern matching values or function calls),
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P}^T)^{n}
\end{align*}
%
The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as shown in Fig. \ref{fig:vectors} (where dashed lines indicate the
composition of the diffent sub-vectors that are calculated separately).

@@

It is possible to provide an overall prediction of the expected number of
constructors when restricting the generation process to only bare data
constructors and pattern matching values.
%
To achieve that, we should stop considering pattern matching values as atomic
constructions and start seeing them as compositions of several data
constructors.
%
In that manner, it is possible to obtain the expected \emph{total} number of
generated data constructors that our generators will produce---regardless if they
are generated on their own, or as part of a pattern matching value.
%
We note this number as $E^\downarrow\![\_]$ and, to calculate it, we only need
to add the expected number of bare constructors that are included within each
pattern matching.
%
For instance, we can calculate the total expected number of constructors |Text|
and |Join| that we will generate by simply expanding the expected number of
generated pattern matching values $simplify\#1$ and $simplify\#2$ into their
corresponding data constructors:
%
{\small
  \begin{align*}
    E^\downarrow\!\!\left[ Text \right]
    = E\left[ Text \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#2 \right] \\
    E^\downarrow\!\!\left[ Join \right]
    = E\left[ Join \right]
    + 1\!\! \cdot\!\! E\!\left[ simplify\#1 \right]
    + 2\! \cdot\!\! E\!\left[ simplify\#2 \right]
  \end{align*}
}
%
Observe that each time we generate a value satisfying the first pattern matching
of the function |simplify|, we add two |Text| and one |Join| data constructors
to our random value.
%
The case of the second pattern matching of |simplify| follows analogously.
%
Note that the overall prediction cannot be applied if we also generate random
values containing function calls, as we cannot predict the output of an
arbitrary function.
